---
title: "ggplot2 Visualization with timechunks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggplot2 Visualization with timechunks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 9,
  fig.height = 4,
  out.width = "100%"
)
library(timechunks)
library(ggplot2)
library(dplyr)
```

## Overview

`timechunks` does **not** provide custom `ggplot2` scales. Instead, it gives
you helper functions to create the label and date columns you pass to `ggplot2`
directly. This is standard tidyverse practice: explicit columns are more
debuggable and composable than hidden scale magic.

There are three common visualization patterns:

| Pattern | When to use | How |
|---|---|---|
| Discrete (equal spacing) | Comparing values across periods | `as_chunk_factor()` |
| Continuous (proportional spacing) | Showing temporal distance | `mid_date()` + `scale_x_date()` |
| Segment (duration-aware) | Showing period lengths | `start_date()` / `end_date()` + `geom_segment()` |

---

## Setup

```{r data}
use_chunk_preset("us_semester")

# Example dataset: enrollment by semester
enrollment <- data.frame(
  semester = time_chunk(c("fa25", "sp26", "su26", "fa26", "sp27", "su27")),
  enrolled = c(8200L, 7100L, 3400L, 8500L, 7300L, 3600L)
)
```

---

## Pattern 1: Discrete scale (equal spacing)

Use `as_chunk_factor()` to convert `time_chunk` to an ordered factor. Each
period gets equal horizontal space.

```{r discrete}
enrollment |>
  mutate(semester_fct = as_chunk_factor(semester, labels = "code")) |>
  ggplot(aes(x = semester_fct, y = enrolled)) +
  geom_col(fill = "steelblue") +
  labs(
    title    = "Enrollment by Semester",
    x        = NULL,
    y        = "Enrolled students"
  ) +
  theme_minimal()
```

Use `labels = "name"` for human-readable labels:

```{r discrete-name}
enrollment |>
  mutate(semester_fct = as_chunk_factor(semester, labels = "name")) |>
  ggplot(aes(x = semester_fct, y = enrolled)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Pattern 2: Continuous scale (proportional spacing)

Use `mid_date()` to place points at the temporal midpoint of each period,
and `format()` for axis labels. Periods with different durations are spaced
proportionally.

```{r continuous}
enrollment |>
  mutate(
    mid = mid_date(semester),
    lbl = format(semester, style = "code")
  ) |>
  ggplot(aes(x = mid, y = enrolled)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue", size = 2) +
  scale_x_date(
    breaks = enrollment |> mutate(mid = mid_date(semester)) |> pull(mid),
    labels = format(enrollment$semester, style = "code")
  ) +
  labs(
    title = "Enrollment Over Time (Proportional Spacing)",
    x     = NULL,
    y     = "Enrolled students"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Pattern 3: Segment visualization (duration-aware) — Recommended

The segment pattern is the most informative: horizontal bars show actual
period duration, so a 5-month Fall looks longer than a 2-month Summer.

```{r segment}
enrollment |>
  mutate(
    x_start = start_date(semester),
    x_end   = end_date(semester),
    x_mid   = mid_date(semester),
    lbl     = format(semester, style = "code")
  ) |>
  ggplot() +
  geom_segment(
    aes(
      x     = x_start,
      xend  = x_end,
      y     = enrolled,
      yend  = enrolled,
      color = lbl
    ),
    linewidth = 4,
    alpha = 0.8
  ) +
  geom_point(aes(x = x_mid, y = enrolled), size = 2) +
  geom_line(aes(x = x_mid, y = enrolled), linetype = "dashed", color = "grey40") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "6 months") +
  labs(
    title  = "Enrollment by Period (Duration-Aware)",
    x      = NULL,
    y      = "Enrolled students",
    color  = "Period"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Grouping by academic year

`chunk_ay()` returns the academic year string, which you can use directly
in `group_by()` or as a facet variable.

```{r ay-facet}
enrollment |>
  mutate(
    ay          = chunk_ay(semester),
    semester_fct = as_chunk_factor(semester, labels = "name")
  ) |>
  ggplot(aes(x = semester_fct, y = enrolled, fill = ay)) +
  geom_col() +
  facet_wrap(~ay, scales = "free_x") +
  labs(
    title = "Enrollment by Academic Year",
    x     = NULL,
    y     = "Enrolled students",
    fill  = "Academic Year"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Arithmetic in plots: building sequences

Use arithmetic to build complete axis ranges programmatically:

```{r sequences}
all_semesters <- seq(time_chunk("fa25"), time_chunk("su27"), by = 1L)

data.frame(
  semester = all_semesters,
  enrolled = c(8200L, 7100L, 3400L, 8500L, 7300L, 3600L)
) |>
  mutate(semester_fct = as_chunk_factor(semester, labels = "code")) |>
  ggplot(aes(x = semester_fct, y = enrolled)) +
  geom_col(fill = "steelblue") +
  labs(title = "Built with seq()", x = NULL, y = "Enrolled") +
  theme_minimal()
```
